---
title: "Analysis of Jail Deaths"
output: pdf_document
---

```{r, setup}
library(tidyverse)
library(lubridate)
library(readr)
library(dplyr)
library(knitr) 
library(tinytex)
```

```{r, import_data}

# import from Micah's local computer
all_jails <- read.csv("C:\\Users\\Micah Clark Moody\\Desktop\\R\\2022-09 Natl Jail Crisis\\Allstatesinsurvey\\all_jails.csv")

```

# 1. Total # of jail deaths in the country since ____

There is not an exact count of people who have died while their physical freedom was constrained by criminal legal systems in the US. 

# 2. 3-5 counties/jurisdictions with the largest # of jail deaths
## 2a. Counties with the most jail deaths in 2019
d2019 = deaths in 2019
```{r, 2a}
all_jails %>%
  slice_max(order_by = d2019, n = 20) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019) %>%
  kable()

```

## 2b. Counties with the most jail deaths from 2008 to 2019

```{r, 2b, include=FALSE}

```

## 2c. Jail deaths by jail population in 2019
d2019 = deaths in 2019
adp2019 = average daily population in 2019
```{r, 2c}
all_jails %>%
  mutate(d_by_adp = d2019 / adp2019) %>%
  slice_max(order_by = d_by_adp, n = 20) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019, d_by_adp) %>%
  kable()

```

## 2d. Jail deaths by jail population from 2008 to 2019

```{r, 2d, include=FALSE}

```

# 3. 3-5 counties/jurisdictions with the lowest # of jail deaths
## 3a. Counties with the least jail deaths in 2019 (not instructive)
d2019 = deaths in 2019

Result: 217 of 523 jails with 0 reported deaths in 2019

```{r, 3a, include=FALSE}
all_jails %>%
  slice_min(order_by = `d2019`)

```

## 3b. Jails in 2019 in the 100 highest dail population with the lowest number of jail deaths
d2019 = deaths in 2019

```{r, 3b}
all_jails %>%
  slice_max(order_by = adp2019, n = 100) %>%
  slice_min(order_by = `d2019`) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019) %>%
  kable()

```

## 3c. Counties with the least jail deaths from 2008 to 2019 (most instructive)

```{r, 3c, include=FALSE}

```

## 3d. Jails in 2019 in the 100 highest dail population with the lowest number of jail deaths by population
d2019 = deaths in 2019
adp2019 = average daily population in 2019
```{r, 3d}
all_jails %>%
  mutate(d_by_adp = d2019 / adp2019) %>%
  slice_max(order_by = adp2019, n = 100) %>%
  slice_min(order_by = d_by_adp, n = 20) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019, d_by_adp) %>%
  kable()

```

## 3e. Jail deaths by jail population from 2008 to 2019

```{r, 3e, include=FALSE}

```

