---
title: "Investigative Memorandum re National Jail Death Crisis"
author: "Micah Clark Moody"
output: github_document
date: 2022-09-21
---

This document has the code corresponding to [this investigative memorandum](https://docs.google.com/document/d/1PwExEtuXhfnc-PM-9RGYx75Xs753CiVJZTHKCIE7_5E/edit#heading=h.1895lafuquc4) on the national jail crisis. Please refer to the memo for background and interpretation of results.

Note: When jail deaths or average daily populations were not recorded in the data set, I replaced the missing value with zero. This will lead to an under count of jail deaths and an undercount of average daily population. While there are few missing values, if you're interested in digging deeper into jails with few incidence of death, it's important to double-check the death incidence were in fact reported.

```{r, setup, echo = FALSE, message = FALSE}
# installing needed packages
library(tidyverse) 
library(lubridate)
library(readr)
library(dplyr)
library(knitr)

# hiding code and messages for all code chunks
opts_chunk$set(echo = FALSE)
opts_chunk$set(message = FALSE)

```

```{r, import_data}

# import from Micah's local computer
all_jails <- read.csv("C:\\Users\\Micah Clark Moody\\Desktop\\R\\2022-09 Natl Jail Crisis\\jail_deaths\\jail_deaths\\reuters_data\\all_jails.csv")

louisiana_jails <- read.csv("C:\\Users\\Micah Clark Moody\\Desktop\\R\\2022-09 Natl Jail Crisis\\jail_deaths\\jail_deaths\\inc_transparency_data\\inc_transparency_data.csv")

d_cols <- c("d2008", "d2009", "d2010", "d2011", "d2012", "d2013", "d2014", "d2015", "d2016", "d2017", "d2018", "d2019")

# creating lists of columns with average daily population 
adp_cols <- c("adp2008", "adp2009", "adp2010", "adp2011", "adp2012", "adp2013", "adp2014", "adp2015", "adp2016", "adp2017", "adp2018", "adp2019")

# creating column names lists
names_total <- c("State", "County", "Jail", "Deaths", "Average Daily Population")
names_by_pop <- c("State", "County", "Jail", "Deaths", "Average Daily Population", "Deaths per Average Daily Population")

# creating varible for total deaths from 2008 to 2019
all_jails$d_total <- all_jails %>%
  select(all_of(d_cols)) %>%
  
  # replacing NA values with 0 which will undercount deaths
  mutate_all(~replace(., is.na(.), 0)) %>%
  rowSums()

# creating variable for mean daily population from 2008 to 2019
all_jails$adp_mean <- all_jails %>%
  select(all_of(adp_cols)) %>%
  
  # replacing NA values with 0 which will undercount population
  mutate_all(~replace(., is.na(.), 0)) %>%
  rowMeans()

# counting missing values
all_jails_d_cols <- all_jails %>%
  select(d_cols)

all_jails$d_NA_count <- rowSums(is.na(all_jails_d_cols))


louisiana_jails %>%
  group_by(Facility) %>%
  summarize(n())
```

# Data limitations

### Smaller facilities
```{r, LA_reuters_comp}

# louisiana totals from Reuters
LA_R_deaths <- all_jails %>%
  filter(state == "Louisiana") %>%
  select(c("d2008", "d2009", "d2010", "d2011", 
           "d2012", "d2013", "d2014", "d2015", "d2016", 
           "d2017", "d2018", "d2019")) %>%
  mutate_all(~replace(., is.na(.), 0)) %>%
  colSums()

LA_R_years <- c(2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019)

LA_R <- data.frame(LA_R_years, LA_R_deaths)

# louisiana totals from incarceration transparency for Reuters data
LA_IC_sample <- louisiana_jails %>%
  filter(Facility %in% c("Caddo Parish Correctional Center", "East Baton Rouge Parish Prison",
                         "Franklin Parish Detention Center", "Jackson Parish Correctional Center - LaSalle",
                         "Jefferson Parish Correctional Center", "Lafayette Parish Correctional Center",
                         "Orleans Parish Prison & TDC", "Ouachita Correctional Center",
                         "Richwood Correctional Center - LaSalle", "St. Tammany Parish Jail"
                         )) %>%
  group_by(Year.Of.Death) %>%
  summarize("Sample Jails, Deaths counted by Incarceration Transparency" = n())

# louisiana totals from incarceration transparency
louisiana_jails %>%
  group_by(Year.Of.Death) %>%
  summarize("All Jails, Deaths" = n()) %>%
  left_join(LA_IC_sample) %>%
  left_join(LA_R, by = c("Year.Of.Death" = "LA_R_years")) %>%
  filter(Year.Of.Death >= 2014, Year.Of.Death <= 2019) %>%
  kable(caption = "Table: Comparing Incarceration Transparency and Reuters Count of Jail Deaths in Louisiana", 
        col.names = c("Year", "All Jail Deaths, IC Count", "Sample Jail Deaths, IC County", "Sample Jail Deaths, Reuters Count"))

```

### Counting missing values

This table counts the number of years with missing death counts by jail. For example, 5 years were missing data for 3 jails.

```{r}
all_jails %>%
  group_by(d_NA_count) %>%
  summarize(count = n()) %>%
  kable(caption = "Table: Counting the incidence of missing data",
        col.names = c("Number of Years with Missing Death Counts", "Number of Jails"))
```

# 1. Total # of jail deaths in the country since ____

See [full memo](https://docs.google.com/document/d/1PwExEtuXhfnc-PM-9RGYx75Xs753CiVJZTHKCIE7_5E/edit#heading=h.1895lafuquc4)

# 2. 3-5 counties/jurisdictions with the largest # of jail deaths

### Table 2a: Highest Jail population

This table shows the 10 jails with the highest average daily population from 2008 to 2019 for reference.

```{r, biggest_jails}
all_jails %>%
  slice_max(order_by = adp_mean, n = 10) %>%
  
  # printing table
  select(state, county, jail, adp2019) %>%
  kable(col.names = c("State", "County", "Jail", "Average Daily Population"))

```

### Table 2b: Highest Jail Deaths in 2019

This table represents the 20 jails with the most reported deaths in 2019.

```{r, 2019_death_max}
all_jails %>%
  slice_max(order_by = d2019, n = 20) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019) %>%
  kable(col.names = names_total)

```

### Table 2c: Highest Jail Deaths from 2008 to 2019

This table represents the 20 jails with the most reported deaths from 2008 to 2019.

```{r, total_death_max}
all_jails %>%
  slice_max(order_by = d_total, n = 20) %>%
  
  # printing table
  select(state, county, jail, d_total, adp_mean) %>%
  kable(col.names = names_total)
```

### Table 2d: Jail deaths by jail population in 2019

This table represents the 20 jails with the highest rate of jail death in 2019. This is calculated by dividing the total deaths by the average daily population. This is to identify small and large jails with high rates of death. 

For example, in this comparison a jail with 1 death and an average daily population of 1000 will have the same death rate as a jail with 10 deaths and an average daily population of 10,000.

```{r, 2019_d_by_adp}
all_jails %>%
  mutate(d_by_adp = d2019 / adp2019) %>%
  slice_max(order_by = d_by_adp, n = 20) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019, d_by_adp) %>%
  kable(col.names = names_by_pop)

```

### Table 2e: Jail deaths by jail population from 2008 to 2019

This table represents the 20 jails with the highest rate of jail death from 2008 to 2019. The methods are identical to Table 2d above.

```{r, d_by_adp_total}
all_jails %>%
  mutate(d_by_adp = d_total / adp_mean) %>%
  slice_max(order_by = adp_mean, n = 20) %>%
  
  # printing table
  select(state, county, jail, d_total, adp_mean, d_by_adp) %>%
  kable(col.names = names_by_pop)

```

# 3. 3-5 counties/jurisdictions with the lowest # of jail deaths

### Table 3a: Jails in the 50 highest jail population with the lowest number of jail deaths in 2019.

This table represents high-population jails with the lowest rates of jail death in 2019.

```{r, min_d}
all_jails %>%
  slice_max(order_by = adp2019, n = 50) %>%
  slice_min(order_by = `d2019`, n = 10) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019) %>%
  kable(col.names = names_total)

```

### Table 3b: Jails in the 50 highest jail population with the lowest number of jail deaths from 2008 to 2019

This table represents high-population jails with the lowest rates of jail death from 2008 to 2019. Excludes jails that did not report data for one or more years.

```{r, 3b}
all_jails %>%
  mutate(d_by_adp = d_total / adp_mean) %>%
  filter(d_NA_count == 0) %>%
  slice_max(order_by = adp2019, n = 100) %>%
  slice_min(order_by = d_by_adp, n = 10) %>%
  
  # printing table
  select(state, county, jail, d_total, adp_mean, d_by_adp) %>%
  kable(col.names = names_by_pop)
```

### Table 3c: 2008 to 2019 high-population jails with lowest death rates

This table represents high-population jails with the lowest rates of jail death from 2008 to 2019.

```{r, 3c}
all_jails %>%
  slice_max(order_by = adp_mean, n = 100) %>%
  slice_min(order_by = d_total, n = 10) %>%
  
  # printing table
  select(state, county, jail, d_total, adp_mean, d_NA_count) %>%
  kable(col.names = c("State", "County", "Jail", "Deaths", "Average Daily Population", "Years with Unreported Data"))
```

### Table 3d: Jails in 2019 in the 50 highest jail populations with the lowest death rates in 2019

This table represents high-population jails with the lowest rates of jail death in 2019.

```{r, 3d}
all_jails %>%
  mutate(d_by_adp = d2019 / adp2019) %>%
  slice_max(order_by = adp2019, n = 100) %>%
  slice_min(order_by = d_by_adp, n = 20) %>%
  
  # printing table
  select(state, county, jail, d2019, adp2019, d_by_adp) %>%
  kable(col.names = names_by_pop)

```

### Table 3e: 2008 to 2019 jails with lowest death rates overall

This table represents high-population jails with the lowest rates of jail death from 2008 to 2019.

```{r, low_d_all_jails}
all_jails %>%
  slice_min(order_by = d_total, n = 30) %>%
  filter(d_NA_count == 0) %>%
  
  # printing table
  select(state, county, jail, d_total, adp_mean) %>%
  kable(col.names = names_total)
```
